---
output: "pdf_document"
editor_options: 
  chunk_output_type: console
urlcolor: blue
fontsize: 12pt
header-includes:
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  comment = "#", 
  cache = FALSE,  
  collapse = TRUE,
  error = TRUE,
  tidy.opts=list(width.cutoff=65)
)
```

# CSI 2300: Intro to Data Science

## In-Class Exercise 09: Exploratory Data Analysis 

**Evelyn Pan**

The data for today's exercises come from the `mowater` library, the `eml`
dataset. This data are about measurements of the properties of water in the Eagle
Mountain Lake reservoir in North Texas.

1. Load the `mowater` library, and then load the `eml` dataset. In the RMarkdown
   document, show the commands you use to do this, but not the
   output of those commands (`message=FALSE` as an option in the code chunk
   header is a good way to do this).
  
```{r, message=FALSE}
library(mowateR)
data(eml)
```

2. Next, inspect the dataset.
   - How many variables are there? **There are 7 variables**
   - How many observations are there? **There are 35532 observations**
   - Are the data already sorted in time order? **Yes, the data is sorted**
   - What do the variables represent, do you think? You may want to consult the
     `help` for the dataset to understand it better, including the units of
     measurement. 
     **Date-Time; Date and two-hour time measurement was taken.** 
     **Depth; Profile depth, measured in meters.**
     **Temp; water temperature, measured in degree Celsius.**
     **DO; Dissolved Oxygen, measured in mg/L.**
     **DOsat; Dissolved Oxygen Saturation, measured as the percentage of DO relative to what the concentration would be in equilibrium with the atmosphere, calcualted as actual/expected.**
     **pH; pH, measured as standard pH from 0 (basic) to 14 (acidic). Cond; Conductivity in water, measured in (micro Siemens per centimeter).**
     
   - What are the ranges of the values? 
```{r}
#View(eml)

max(eml$Date.Time) - min(eml$Date.Time)
max(eml$Depth) - min(eml$Depth)
max(eml$Temp) - min(eml$Temp)
max(eml$DO) - min(eml$DO)
max(eml$DOsat) - min(eml$DOsat)
max(eml$pH) - min(eml$pH)

```

  - Are there any missing values (NA values)? **No**

3. The size of our dataset is rather large for easy (and fast) manipulation. 
   Pare it down by creating a new data frame with only the last $500$ observations in it. This is
   throwing away a lot of information, and we should be careful any time we do
   this. But if we try to use all the data, it may be too time-consuming for an
   in-class exercise.
   
```{r}
new_eml = tail(eml, 500) 
```



4. Now that the data are of a manageable size for visualizing it, plot all of
   the variables against each other. Try calling `plot(eml_small)`,  but replace
   `eml_small` with the name of the *small* data frame you just created.  This creates a matrix of pairwise scatterplots.  
   (If this takes a very long time, you may have made the mistake of trying to
   plot the original dataset, which is quite large for this task.)
   
```{r}
plot(new_eml)
```


5. Look at the plot and consider which pairs of variables appear to be linearly
   related. Choose `DO` and `pH`.  Make a scatter plot of `DO` on the
   y-axis and `pH` on the x-axis.  Which one are you thinking of as the
   independent variable, and which one are you thinking of as the dependent
   variable?
```{r}
plot(new_eml$DO, new_eml$pH,
     xlab = "pH Levels",
     ylab = "Dissolved Oxygen Levels")
```
The x axis is always the independent variable. Thus, the pH Levels depend on the acidity (pH levels) of the water. 


6. Run a linear regression by calling `lm(DO ~ pH, data=eml_small)`
   where again `eml_small` is the small data frame.  Plot the regression line on
   the scatter plot using `abline` on the model that `lm` returns. If the line
   doesn't appear to follow the data, you may have switched the variables (for
   the plot versus the linear regression).
   
```{r}
reg = lm(DO ~ pH, data=new_eml)
plot(new_eml$pH, new_eml$DO,
     ylab = "pH Levels",
     xlab = "Dissolved Oxygen Levels")
abline(reg, col = "navyblue")
``` 


7. Investigate the coefficients and summary statistics of the model that `lm`
   gave you. Comment on the coefficient values, significance levels of the
   coefficients (are they significantly different from $0$), and $R^2$ values.
   Does there appear to be a linear relationship between these two variables?
   
```{r}
summary(reg)
``` 
Coefficient Values: The estimated 
Significance levels of coefficients: 
Linear relationship? 

8. Use the model to make a prediction of the dependent variable for when the
   variable `pH` is $7.7$. You can do this by direct computation if you want
   (rather than using any specialty R command).  Hand-check your work on your
   plot. Is the predicted value close to the plotted data? **The predicted value of 7.7 DO level is close to the pH level on the y-axis of 2.8677**
```{r}

13.182 * 7.7 - 98.6337

```


9. Do you trust this model to make a prediction of `DO` for a `pH` value of 3?
   Explain your answer.
**No, because if you have a value that's out of the x range of 7.4 - 7.9, you can't extrapolate the data that's out of the x range.**


10. Extra credit: repeat steps 5-7, but for a different pair of variables than
   the pair you were just working with. Compare the models; does it appear
   that one pair of variables is more strongly linearly related than the other
   pair? Note, do not choose the pair `DO` and `DOsat` as they are the same
   variable but measured in different units. **Since it's less than 1, it proves evidence that it is less than zero.**
```{r}
plot(new_eml$Temp, new_eml$pH,
     xlab = "Temp",
     ylab = "pH Levels")

reg = lm(Temp ~ pH, data=new_eml)
plot(new_eml$pH, new_eml$Temp,
     ylab = "Temp",
     xlab = "pH Levels")
abline(reg, col = "hotpink")

summary(reg)

```

